plyranges: a grammar for transforming genomics data

### 2018-04-26 paper.Rmd
rework of algebra part

The Bioconductor infrastructure models the genomic data and operations from the
perspective of the power user, one who understands and wants to take
advantage of the subtle differences in data types. This design has
enabled the development of sophisticated tools, as evidenced by the
hundreds of packages depending on the framework. Unfortunately, the
myriad of data structures have overlapping purposes and important but
obscure differences in behaviour that often confuse the typical end
user.

@@

# Design and Implementation

The `plyranges` DSL is built on the most general Bioconductor data structure, 
GRanges, which is capable of
representing all types of genomic data at a semantic level that
roughly matches the intuition of most users.  A GRanges is essentially
a table, with columns for the chromosome, start and end coordinates,
and the strand, along with an arbitrary set of additional columns,
consisting of measurements or metadata specific to the data type or
experiment. For example, a GRanges can represent a gene with
its chromosomal coordinates, exonic structure, and a count column
summarized from an RNA-seq experiment.

@@

## Genomic Relational Algebra

The `plyranges` DSL defines an expressive algebra for performing genomic
arithmetic with and between GRanges objects. Often in analyses of
genomic regions we are required to make adjustments to the start, end,
and width columns. This is supported directly in `plyranges` via extending
the `mutate()` function in `dplyr` (table \ref{tab:grammar}). 
However, often adjustments
need to be relative: there is a mutual dependence between the start, end and
width columns over a genomic region. Changing the start,column needs to change
either the width or the end to preserve integrity of the object. The GRanges 
object by design, causes a user to think of the range as a pair of endpoints, 
and hence we have developed operators that perform coordinate transformations with 
the notion of an 'anchoring' combinator.

For width modification, we can stretch the range by a fixed amount or
resize the width of a range. The 'anchoring' combinator preserves one of the 
endpoints or the  midpoint of a range when altering performing
width modification. It can also be used to preserve coordinates over strand, 
for example, anchoring over the three prime end of a genomic region, will 
preserve the start on the negative strand and end on the positive strand. 
Anchoring the width would not make sense, because it is not a point that can be
intuitively anchored and stretching and resizing obviously modify the width, 
so it would not apply to those.

Transformations that are width invariant are designed to be explicit
about the direction to modify and are performed via the shift operator. 
For GRanges objects strand modulates direction and we distinguish left/right 
(chromosomal coordinates) from the direction of transcription. To generate
flanking regions, we use the flank operator which is also explicit about the
direction of the input range.

Binary operations between two ranges mostly align with set
operations that treat ranges as sets of integers. These are
parallel/vectorized operations between two vectors/tables of
ranges, and are not to be confused with the merging 
operations defined in table \ref{tab:grammar}. We make the binary nature of 
these operators
by having in-fix functions in R for the parallel union, intersect and
asymmetric set difference. There are also operations that relate to the 
relative positions of two ranges such as the 'between' operator which
returns the range in the gap between two ranges and the 'span' operator 
a range that spans both ranges. The binary operations can be expressed within
the arithmetic, restriction or aggregation functions in table \ref{tab:grammar}.

Our algebra recasts the actions of finding overlaps or nearest neighbours 
between two genomic regions as a relational join operator. The join operator 
acts on two GRanges object, a query and a subject. The join operator is 
relational in the sense that metadata from the query and subject ranges is 
retained in the joined range.  The commonality between
all join operators in the `plyranges` DSL is to use the index of where in
the query range the subject range 'hits' the query range as a primary key 
(figure \ref{fig:olaps-fig}). The notion of 'hits' and hence
the resulting key is determined by the type of join (preceding, nearest, follows
overlaps).

@@

We extend this idea by introducing three types of overlap joins: inner,
intersect and left (figure \ref{fig:olaps-fig}). The inner join considers a
hit to be any overlap between subject and query and expands the query range to 
have length equal to the  number of hits. The intersect join uses the same 
definition of a hit as the inner join but returns a range where the start and end coordinates
are the intersect of coordinates in the subject that hit the query. Finally,
the overlap left join is akin to left outer join in Cobb's relational algebra:
it performs an overlap inner join but also returns all query ranges that
are not hit by the subject.

The behaviour of a join operator can be further altered with additional
suffixes by restricting or expanding how a subject 'hits' a query. For example,
for overlap joins we can use suffixes to encapsulate Allan's Interval Algebra:
to consider subject 'hits' that are entirely 'within' a query range, the 'within'
prefix is used. We also use the 'directed' suffix to explicitly include 
notions of strandedness in our algebra.

>>>>

The Bioconductor infrastructure models the genomic data and operations from the
perspective of the power user, one who understands and wants to take
advantage of the subtle differences in data types. This design has
enabled the development of sophisticated tools, as evidenced by the
hundreds of packages depending on the framework. Unfortunately, the
myriad of data structures have overlapping purposes and important but
obscure differences in behavior that often confuse the typical end
user.

@@

# Genomic relational algebra

## Data model

The `plyranges` DSL is built on the core Bioconductor data structure
GRanges, which is capable of
representing all types of genomic data at a semantic level that
roughly matches the intuition of most users.  A GRanges is essentially
a table, with columns for the chromosome, start and end coordinates,
and the strand, along with an arbitrary set of additional columns,
consisting of measurements or metadata specific to the data type or
experiment. For example, a GRanges can represent a gene with
its chromosomal coordinates, exonic structure, and a count column
summarized from an RNA-seq experiment.

@@

## Algebraic operations

The `plyranges` DSL defines an expressive algebra for performing
genomic operations with and between GRanges objects (see table
\ref{tab:grammar}). [ML: add a couple overview sentences indicating the
classes of operations like arithmetic, merging, aggregation]

GRanges describes the within-sequence coordinates of a range by its
_start_, _end_ and _width_. Those three variables are mutually
dependent and partially redundant, so direct manipulation of them is
problematic. For example, changing the _width_ column needs to change
either the _start_, _end_ or both to preserve integrity of the
object. We introduce the _anchor_ combinator to disambiguate these
adjustments. Supported anchor points include the start, end and
midpoint, as well as the 3' and 5' ends for strand-directed
ranges. For example, if we anchor the start, then setting the width
will adjust the end while leaving the start stationary.

[ML: where does the word 'combinator' come from here? maybe 'modifier'
 is better? that's what the table uses]

The algebra also defines conveniences for relative coordinate
adjustments: _shift_ (unanchored adjustment to both start and end) and
_stretch_ (anchored adjustment of width). We can perform any relative
adjustment by some combination of those two operations.  The _stretch_
operation requires an anchor and assumes the midpoint by
default. Since _shift_ is unanchored, the user specifies a suffix for
indicating the direction: left/right or, for stranded features,
upstream/downstream. For example, _shift\_right_ shifts a range to the
right.

The _flank_ operation generates new ranges that are adjacent to
existing ones. This is useful, for example, when generating upstream
promoter regions for genes. Analogous to _shift_, a suffix indicates
the side of the input range to flank.

As with other genomic grammars, we define set operations that treat
ranges as sets of integers, including _intersect_, _union_,
_difference_, and _complement_. There are two sets of these: parallel
and merging. The parallel operations map to infix operators, which we
surround with _%_ symbols in accordance with R syntax rules.  For
example, the parallel intersection (_x %intersect% y_) finds the
intersecting range between _xi_ and _yi_ for _i_ in _1...n_, where _n_
is the length of both _x_ and _y_. In constrast, the merging
intersection (_intersect\_ranges(x, y)_) returns a new set of disjoint
ranges representing whereever there was overlap between a range in _x_
and a range in _y_. We use the infix syntax for the parallel
operations, since it is the conventional syntax for parallel, binary
operations in R. Finding the parallel union will fail when two ranges
have a gap, so we introduce a _span_ operator that takes the union
while filling any gap. The _complement_ operation is unique in that it
is unary. It finds the regions not covered by any of the ranges in a
single set. Closely related is the _between_ parallel operation, which
finds the gap separating _xi_ and _yi_. The binary operations are
callable from within the arithmetic, restriction and aggregation
expressions.

Our algebra recasts finding overlaps or nearest neighbours between two
genomic regions as variants of the relational join operator. A join
acts on two GRanges objects, a query and a subject. The join operator
is relational in the sense that metadata from the query and subject
ranges is retained in the joined range.  All join operators in the
`plyranges` DSL generate a set of hits based on overlap or proximity
of ranges and use those hits to merge the two datasets in different
ways. There are four supported matching algorithms: _overlap_,
_nearest_, _precede_, and _follow_. We can further restrict the
matching by whether the query is completely _within_ the subject, and
adding the _directed_ suffix ensures that matching ranges have the
same direction (strand).

For merging based on the hits, we have three modes: _inner_,
_intersect_ and _left_. The _inner_ overlap join is similar to the
conventional inner join in that there is a row in the result for every
match. A major difference is that the matching is not by identity, so
we have to choose one of the ranges from each pair. We always choose
the left range. The _intersect_ join uses the intersection instead of
the left range. Finally, the overlap _left_ join is akin to left outer
join in Cobb's relational algebra: it performs an overlap inner join
but also returns all query ranges that are not hit by the subject.

@@

[ add a simple paragraph about how we can filter, sort, group and
  aggregate just like other tabular structures ]
